---
title: "Spring 2025 Homework Modeling Exercises Weeks 1 & 3"
author: "Group 2: group names, University of Chicago"
geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
date: "03-31-2026"
output: pdf_document
header-includes:
  - \usepackage{caption}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{amsmath}
  - \usepackage{enumitem}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.pos = "H",
	message = FALSE,
	warning = FALSE,
	out.extra = ""
)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)

library(data.table)
library(dplyr)
library(tidyr)
library(MASS)
library(ggplot2)
library(ggdensity)
library(plotly)
library(magrittr)
```

## Download data 

3. Set RStudio’s working directory to the folder containing the datasets for this homework and
open the PJM data set (“data_PJM.csv”), as below. This dataset is constructed from hourly
pricing data from PJM’s Data Miner 2 and includes both day-ahead (“DA”) and real-time(“RT”) locational marginal pricing (“LMP”) data.

```{r q3}
data_PJM <- fread('PJM Data.csv')
```

4. Load the following data set, which combines pricing and weather data for Chicago, Milwaukee, and Corpus Christi.

```{r q4}
data_Chicago <- fread('Data Chicago.csv')
data_Milwaukee <- fread('Data Milwaukee.csv')
data_CorpusChristi<- fread('Data Corpus Christi.csv')
data_Austin <- fread('Data Austin.csv')
```

## III. Data Familiarization

5. Let’s explore the PJM data! The following code identifies the fields of the dataset

```{r q5}
names(data_PJM)
```

6. To see the first few observations of the data, run the code below. What is the unit ofobservation?

```{r q6}
head(data_PJM)
```