---
title: "Homework 2, Business 33320, University of Chicago Booth School of Business \n Due Week 4 (April 14, 2026)"
author: "Group 2: Bailey Meche, Eliott Raia, University of Chicago"
geometry: "left=1cm,right=1cm,top=1cm,bottom=1.5cm"
date: "04-13-2026"
output: pdf_document
header-includes:
  - \usepackage{caption}
  - \usepackage{float}
  - \usepackage{xcolor}
  - \usepackage{amsmath}
  - \usepackage{enumitem}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
	echo = TRUE,
	fig.pos = "H",
	message = FALSE,
	warning = FALSE,
	out.extra = ""
)
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)
options(scipen = 999)

library(data.table)
library(dplyr)
library(tidyr)
library(MASS)
library(ggplot2)
library(ggdensity)
library(plotly)
library(magrittr)
# wd <- "C:/Users/Owner/Box/spring26/GridFinancing/hw1/Homework1&3 Files from TZ-selected" # set your working directory here 
# setwd(wd) 
```

## 1. 


## 2. 
Download the data for February 2025 for the MISO LMP’s from the MISO site. The
files are given ashttps://docs.misoenergy.org/marketreports/YYYYMMDD_rt_lmp_final.csvwhere YYYYMMDD stands for the year, month, and day.

### a. For each day, find the mean, maximum, minimum and standard deviation of the LMP’s across all locations for each hour of each day. Group the days into two groups: weekdays and weekends. Within each group, find the averages of the mean, maximum, minimum, and standard deviation of the LMP for each hour and then plot these values for each hour (i.e., for both weekdays and weekends, you should have 4 plots, one for each of these LMP statistics with the LMP metric on  the y-axis and hours on the x-axis).

### b. Describe your observations of these LMP metrics as a function of hour of day and day of week (weekday or weekend) and what implications these may have for acquiring financial transmission rights (a financial instrument which pays the difference in LMP’s between two nodes in the transmission network) for different days of the week and times of day.

```{r q2a}
# code 
```


